Comparing fluorometric methods ( in vivo vs extracted) for cyanoHAB monitoring in six Rhode Island ponds

### 2024-04-05 manuscript/manuscript.qmd

---
title: "Comparing fluorometric methods ( in vivo vs extracted) for cyanoHAB monitoring in six Rhode Island ponds"
author: 
  - name: Stephen D. Shivers
    orcid: 0000-0003-0096-7433
    email: shivers.stephen@epa.gov
    corresponding: true
    affiliation:
      - id: epa
        name: US Environmental Protection Agency, Atlantic Coastal Environmental Sciences Division
        city: Narragansett
        state: RI
        zip: 02882
  - name: Jeffrey W. Hollister
    orcid: 0000-0002-9254-9740
    affiliation:
      - ref: epa
  - name: Sophie Fournier
    affiliation:
      - id: orise
        name: Oak Ridge Institute for Science and Education
        city: Narragansett
        state: RI
        zip: 02882
  - name: Jakob Stankoski
    affiliation:
      - id: orau
        name: Oak Ridge Associated Universities
        city: Narragansett
        state: RI
        zip: 02882
  - name: Betty J. Kreakie
    orcid: 0000-0002-4982-7610
    affiliation:
      - ref: epa
abstract: "Harmful algal blooms caused by cyanobacteria (cyanoHABs) are detrimental to human and environmental health and can be difficult to monitor without specialized training and equipment.  A variety of instruments have been developed to measure cyanoHAB indicators (i.e. chlorophyll *a* or phycocyanin) that do not require advanced laboratory processes (e.g. pigment extraction).  We compared measurements from five in vivo fluorometers (Turner Trilogy in-vivo module, Turner Fluorosense, Turner Cyanofluor, bbe AlgaeTorch, and bbe Phycoprobe) to results from solvent-based extractions for chlorophyll *a* and phycocyanin at six different waterbodies in Rhode Island.  We found a strong relationship between extracted phycocyanin and  in vivo fluorometers (R^2^ ranging from 0.78-0.96).  We found less consistency between  in vivo measurements of chlorophyll *a* and the extracted results (R^2^ between 0.34 and 0.82).  Some variability in the chlorophyll *a* results can be explained by differences in the phytoplankton community at the different sampling locations.  Phycocyanin results from  in vivo fluorometry were also strongly correlated with cell counts, which implies that phycocyanin measurements from these instruments can be a good proxy for cell counts.  Many federal, state, and local entities use cell counts of cyanobacteria to determine when to issue health or contact advisories for waterbodies.  Producing accurate cell counts requires highly specialized training/equipment, processing time, and counts can vary greatly between technicians.  The results from this study encourage further research using in vivo fluorometry and phycocyanin for cyanoHAB monitoring efforts and when making decisions regarding public health."
format: 
    docx:
      reference-doc: manuscript_template.docx
bibliography: manuscript.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  cache = TRUE,
  tidy = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.height = 4.8,
  fig.width = 6,
  dpi = 300
)
```

```{r analysis, include=FALSE}
source(here::here("R/analysis.R")) 
```

@@

```{r chl_summary, results='asis', echo=FALSE, message=FALSE}
#| label: tbl-chlsummary
#| tbl-cap: "Chlorophyll summary data by waterbody and instrument."

@@

```{r phyco_summary, results='asis', echo=FALSE, message=FALSE}
#| label: tbl-phycosummary
#| tbl-cap: "Phycocyanin summary data by waterbody and instrument."

@@

Cyanophytes were the dominant taxa by relative biovolume at Curran (0.75), Mashapaug (0.87), and Warwick (0.89); Chrysophytes at Barber (0.39) and Yawgoo (0.44); and Bacillariophytes at Indian (0.41)(@fig-cellsml).  Total biovolume concentrations were highest at Mashapaug (2.12x10^7^), Warwick (1.44x10^7^), and Curran (3.22x10^6^)(@fig-bv). Microcystis was the dominant genus at Mashapaug and Warwick, while Chrysosporum was dominant at Curran (@tbl-supp).

>>>>

---
title: "Comparing fluorometric methods ( in vivo vs extracted) for cyanoHAB monitoring in six Rhode Island ponds"
author: 
  - name: Stephen D. Shivers
    orcid: 0000-0003-0096-7433
    email: shivers.stephen@epa.gov
    corresponding: true
    affiliation:
      - id: epa
        name: US Environmental Protection Agency, Atlantic Coastal Environmental Sciences Division
        city: Narragansett
        state: RI
        zip: 02882
  - name: Jeffrey W. Hollister
    orcid: 0000-0002-9254-9740
    affiliation:
      - ref: epa
  - name: Sophie Fournier
    affiliation:
      - id: orise
        name: Oak Ridge Institute for Science and Education
        city: Narragansett
        state: RI
        zip: 02882
  - name: Jakob Stankoski
    affiliation:
      - id: orau
        name: Oak Ridge Associated Universities
        city: Narragansett
        state: RI
        zip: 02882
  - name: Betty J. Kreakie
    orcid: 0000-0002-4982-7610
    affiliation:
      - ref: epa
abstract: "Harmful algal blooms caused by cyanobacteria (cyanoHABs) are detrimental to human and environmental health and can be difficult to monitor without specialized training and equipment.  A variety of instruments have been developed to measure cyanoHAB indicators (i.e. chlorophyll *a* or phycocyanin) that do not require advanced laboratory processes (e.g. pigment extraction).  We compared measurements from five in vivo fluorometers (Turner Trilogy in-vivo module, Turner Fluorosense, Turner Cyanofluor, bbe AlgaeTorch, and bbe Phycoprobe) to results from solvent-based extractions for chlorophyll *a* and phycocyanin at six different waterbodies in Rhode Island.  We found a strong relationship between extracted phycocyanin and  in vivo fluorometers (R^2^ ranging from 0.78-0.96).  We found less consistency between  in vivo measurements of chlorophyll *a* and the extracted results (R^2^ between 0.34 and 0.82).  Some variability in the chlorophyll *a* results can be explained by differences in the phytoplankton community at the different sampling locations.  Phycocyanin results from  in vivo fluorometry were also strongly correlated with cell counts, which implies that phycocyanin measurements from these instruments can be a good proxy for cell counts.  Many federal, state, and local entities use cell counts of cyanobacteria to determine when to issue health or contact advisories for waterbodies.  Producing accurate cell counts requires highly specialized training/equipment, processing time, and counts can vary greatly between technicians.  The results from this study encourage further research using in vivo fluorometry and phycocyanin for cyanoHAB monitoring efforts and when making decisions regarding public health."
format: 
    docx:
      reference-doc: manuscript_template.docx
bibliography: manuscript.bib
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  tidy = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.height = 4.8,
  fig.width = 6,
  dpi = 300
)
```

```{r analysis, include=FALSE, cache=FALSE}
source(here::here("R/analysis.R")) 
```

@@

```{r chl_summary, results='asis', echo=FALSE, message=FALSE}
#| label: tbl-chlsummary
#| tbl-cap: "Extracted chlorophyll summary data by waterbody and instrument."

@@

```{r phyco_summary, results='asis', echo=FALSE, message=FALSE}
#| label: tbl-phycosummary
#| tbl-cap: "Extracted phycocyanin summary data by waterbody and instrument."

@@

<!-- phycotech data is in data/phycotech_data_summ.csv" -->

Cyanophytes were the dominant taxa by natural units / mL at Curran (???), Mashapaug (0.87), and Warwick (0.89); Chrysophytes at Barber (0.39) and Yawgoo (0.44); and Bacillariophytes at Indian (0.41)(@fig-cellsml).  Total biovolume concentrations were highest at Mashapaug (2.12x10^7^), Warwick (1.44x10^7^), and Curran (3.22x10^6^). Microcystis was the dominant genus at Mashapaug and Warwick, while Chrysosporum was dominant at Curran (@tbl-supp).
